### 2024-06-15 research-methods.qmd
20240615 - mediation/moderation

In an *experiment*, there are one or more things that we manipulate to see how the manipulation influences the process of interest.
The variable that we manipulate is the *independent variable*.
By contast, the *dependent variable* is the variable that we evaluate to determine whether it was influenced by the manipulation (i.e., by the independent variable).

@@

In a case study, we assess a small sample of individuals (commonly only one person or a few people), often with rich qualitative information.
Themes may be coded from the qualitative information, which may help inform inferences about whether some process may have played a role in influencing the outcome of interest.

For instance, to use a case study to evalute the possibilty that players who drink sports drinks perform better than players who drink only water, we could conduct an in-depth interview with a player.
In the interview, we might ask the player how they performed in games with versus without a sports drink and have them discuss whether they believe the sports drink improved their performance (and if so, how).
Then, based on the player's responses, we might code the responses to extract themes and to make a qualitative judgement of whether or not the player likely performed better during games in which they had a sports drinl

@@

@@

@@

A research design can also be within-subject, between-subject, or both.

@@

@@

>>>>

In an *experiment*, there are one or more things (i.e., variables) that we manipulate to see how the manipulation influences the process of interest.
The variable that we manipulate is the *independent variable*.
By contast, the *dependent variable* is the variable that we evaluate to determine whether it was influenced by the manipulation (i.e., by the independent variable).Besides the independent and dependent variables, the researcher attempts to hold everything else constant through processes including standardization and random assignment.
*Standardization* involves using the same procedures to assess each participant, so that scores can be fairly compared across participants (and groups).
Random assignment involves randomly assigning participants to conditions of the independent variable, so the people in each condition are comparable and do not differ systematically.

@@

In a case study, we assess a small sample of individuals (commonly only one person or a few people), often with rich qualitative information.
Themes may be coded from the qualitative information, which may help inform inferences about whether some process may have played a role in influencing the outcome of interest.
The inferences are then drawn in a subjective, qualitative way.
Testimonials and anecdotes are examples are case studies.

For instance, to use a case study to evalute the possibilty that players who drink sports drinks perform better than players who drink only water, we could conduct an in-depth interview with a player.
In the interview, we might ask the player how they performed in games with versus without a sports drink and have them discuss whether they believe the sports drink improved their performance (and if so, how).
Then, based on the player's responses, we might code the responses to extract themes and to make a qualitative judgement of whether or not the player likely performed better during games in which they had a sports drink.

@@

Cross-sectional studies are more common than longitudinal studies because cross-sectional studies are less costly and time-consuming.
They can provide a helpful starting point to test findings more rigorously in subsequent longitudinal studies.

@@

Longitudinal studies are less common than cross-sectional studies because longitudinal studies are more costly and time-consuming.
Nevertheless, longitudinal studies can allow us test our hypotheses more rigorously, because they can allow us to test whether changes in the predictor/indepdnent variable leads to changes in the outcome/dependent variable.
Thus, compared to cross-sectional studies, longitudinal studies can provide greater confidence in causal inferences.

@@

A research design can also be within-subject, between-subject, or both.
A study can involve both within-subject and between-subject comparisons if one predictor/independent variable is within-subject and another predictor/independent variable is between-subject.

@@

### Tradeoffs Between Internal and External Validity {#sec-internalExternalValidityTradeoffs}

There is a tradeoff between [internal](#sec-internalValidity) and [external](#sec-externalValidity) validity—a single research design cannot have both high [internal](#sec-internalValidity) and high [external validity](#sec-externalValidity).
Each study and design has weaknesses.
Some research designs are better suited for making causal inferences, whereas other designs tend to be better suited for making inferences that generalize to the real world.
The research design that is best suited to making causal inferences is an [experiment](#sec-experiment) because it is the design in which the researcher has the greatest control over the variables.
Thus, [experiments](#sec-experiment) tend to have higher [internal validity](#sec-internalValidity) than other research designs.
However, by manipulating one variable and holding everything else constant, the research takes place in a very standardized fashion that can become like studying a process in a vacuum.
So, even if a process is theoretically causal in a vacuum, it may act differently in the real world when it interacts with other processes.

[Correlational designs](#sec-correlationalStudy) have greater capacity for [external validity](#sec-externalValidity) than [experimental designs](#sec-experiment) because the participants can be observed in their natural environments to evaluate how variables are related in the real world.
However, the greater [external validity](#sec-externalValidity) comes at a cost of lower [internal validity](#sec-internalValidity).
[Correlational designs](#sec-correlationalStudy) are not well-positioned to make causal inferences.
As the maxim goes, "correlation does not mean causation"—just because two variables are associated does not necessarily mean that they are causally related.
There could be third variables (i.e., confounds) that act as a common cause of both the predictor and outcome variables, thus explaining why they are correlated.

For instance, it may not be that sport drink consumption per se influences player performance; rather, it may be that players who are more intelligent or have more financial resources tend to drink more sports drinks and also tend to perform better.
In this case, intelligence or financial resources may be a confound that influences both sports drink consumption and player performance, but sports drink consumptions—though correlated with player performance—does not influence player performance.
Or, the variables could be associated due to the opposite direction of effect (i.e., the outcome variable may influence the predictor variable).
For instance, greater performance may lead players to drink more sports drink (rather than the reverse).
[Correlational studies](#sec-correlationalStudy) can account for potential confounds using *covariates* or for the reverse direction of effect using longitudinal designs, but the researcher has less control over the variables than in an [experiment](#sec-experiment).

As the [internal validity](#sec-internalValidity) of a study's design increases, its [external validity](#sec-externalValidity) tends to decrease.
The greater control we have over variables (and, therefore, have greater confidence about causal inferences), the lower the likelihood that the findings reflect what happens in the real world because it is studying things in a metaphorical vacuum.
Because no single research design can have both high [internal](#sec-internalValidity) and [external](#sec-externalValidity) validity, scientific inquiry needs a combination of many different research designs so we can be more confident in our inferences—[experimental designs](#sec-experiment) for making causal inferences and [correlational designs](#sec-correlationalStudy) for making inferences that are more likely to reflect the real world.

[Case studies](#sec-caseStudy), because they have smaller sample sizes and inferences drawn in a subjective, qualitative way, tend to have lower [external validity](#sec-externalValidity) than both [experimental](#sec-experiment) and [correlational](#sec-correlationalStudy) studies.
[Case studies](#sec-caseStudy) also tend to have lower [internal validity](#sec-internalValidity) because they have less control over variables, and thus fail to remove the possibility of illusory correlations, potential confounds, or the reverse direction of effect.
Thus, [case studies](#sec-caseStudy) are among the weakest forms of evidence.
Nevertheless, case studies can still be useful for generating hypotheses that can then be tested empirically with a larger sample in [experimental](#sec-experiment) or [correlational](#sec-correlationalStudy) studies.

@@

## Mediation vs Moderation {#sec-mediationModeration}

Both types of effects involve (at least) three variables:

1. An independent/predictor variable, which will be labeled as `X`.
1. A dependent/outcome variable, which will be labeled as `Y`.
1. The mediator or moderator variable, which will be labeled as `M`.

![Mediation Versus Moderation Mnemonic](images/mediationModerationMnemonic.png)

### Mediation {#mediation}

#### Overview {#mediationOverview}

**Mediation** is a <u>causal chain of events</u>, where one variable (a mediator variable) at least partially <u>explains (or accounts for) the association between two other variables</u> (the predictor variable and the outcome variable).
In mediation, a predictor (`X`) leads to a mediator (`M`), which leads to an outcome (`Y`).
Mediation answers the question of, "<u>**Why (or how)** does `X` influence `Y`?</u> A <u>mediator</u> (`M`) is a variable that helps explain the assocation between two other variables, and it answers the question of why/how `X` influences `Y`.
That is, the mediator is the variable that helps explain how/why `X` is related to `Y`.
In other words, you can think of the mediator as the <u>mechanism</u> that helps explain why `X` has an impact on `Y`.
The association between `X` and `Y` <u>gets smaller</u> when accounting for `M`.
Visually this can be written as:

![Mediation](images/mediation.png)

where `X` is causing `M`, which in turn is causing `Y`.
In other words, `X` leads to `M`, and `M` leads to `Y`.

For instance, if we determine that consuming sports drinks improves player performance, we may want to know how/why.
That is, what is the mechanism that leads consumption of sports drinks <u>to improve</u> player performance?
We might hypothesize that consumption of sports drink helps increase a player's hydration, which in turn will improve the player's performance.
In this case, increased hydration <u>mediates</u> (i.e., helps explain or account for) the effect of the sports drink consumption on improved player performance.

Question: <u>Why/how</u> does sports drink consumpion lead players to perform better?\
Answer: increased hydration

As a picture, we can draw this assocation as:

[INSERT]

#### Types of Mediation {#mediationTypes}

##### Full Mediation {#fullMediation}

When one mechanism fully accounts for the effect of the predictor variable on the outcome variable, this is known as **full mediation**, as depicted below:

![Full Mediation](images/mediation.png)

##### Partial Mediation {#partialMediation}

When a single process partially—but does not fully—accounts for the effect of the predictor variable on the outcome variable; this is known as **partial mediation** and is depicted below:

![Partial Mediation](images/mediationPartial.png)

##### Multiple Mediators {#multipleMediators}

In addition, there can be multiple mediators/mechanisms that account for the effect of a predictor variable on an outcome variable, as depicted below:

![Multiple Mediators](images/mediationMultiple.png)

### Moderation (i.e., Interaction) {#moderation}

#### Overview {#moderationOverview}

**Moderation** (sometimes called an "interaction"), on the other hand, occurs when there is a variable or condition (`M`; called a "moderator") that <u>changes the assocation</u> between `X` and `Y`.
That is, the effect of the predictor variable on the outcome variable differs at different levels of the moderator variable.
In these cases, `X` and `M` <u>work together</u> to have an effect on `Y`; here `X` <u>does not have a direct effect</u> on `M`.
Moderation answers the question of, "<u>**For whom** does `X` influence `Y`?</u>"
If `X` influences `Y` more strongly for some people or in some circumstances, we would say that there is an interaction such that the <u>effect of `X` on `Y` depends on `M`</u>, as depicted below:

![Moderation](images/moderation.png)

For example, if the effect of consuming sports drinks on player performance differs for Quarterbacks and Wide Receivers, the interaction could be depicted below:

[INSERT]

An interaction can be identified visually by <u>non-parallel lines</u> at different levels of the moderator.
